Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-07 formally.tex
final sec 4

Before giving the rules for these judgements, let us state two key metatheorems. These metatheorems deeply inform the design of the rules. Their proofs, which are mechanized in Agda, are by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules, starting in Sec. \ref{sec:action-subsumption}.

% The type action judgement appears as a premise in the rules that handle Z-expressions where the cursor is inside a . 
% \[\begin{array}{ll}
% %\textbf{Judgement Form} & \textbf{Description}\\
% & 
% \end{array}\]
% \[
% \begin{array}{l}
% \fbox{$\performTyp{\ztau}{\alpha}{\ztau'}$} ~~ \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
% \fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
% \quad \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
% \fbox{$\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
% \quad \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
% \quad \text{against $\htau$}
% \end{array}\]
\subsubsection{Sensibility}
The expression action judgements obey the following \emph{action sensibility} theorem:
\begin{theorem}[Action Sensibility]\label{thrm:actsafe} Both of the following hold:\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ 
   then
  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{enumerate}
\end{theorem}
\noindent In other words, if an edit state (i.e. a Z-expression) is statically meaningful, i.e. its cursor erasure is well-typed, then performing an action on it leaves the resulting edit state statically meaningful. In particular, the first clause of Theorem \ref{thrm:actsafe} establishes that when an action is performed on an edit state whose cursor erasure synthesizes an H-type, the result is an edit state whose cursor erasure also synthesizes some (possibly different) H-type. The second clause establishes that when an action is performed using the analytic action judgement on an edit state whose cursor erasure analyzes against some H-type, the result is a Z-expression whose cursor erasure also analyzes against the same H-type. % Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.

@@

\subsubsection{Determinism}
Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
\begin{theorem}[Action Determinism] All of the following hold:
\label{thrm:actdet}
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
  $\zexp' = \zexp''$ and $\htau' = \htau''$.
% \item If all of

%   \begin{quote}
%     \begin{enumerate}
%     \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
%     \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
%     \item $\tcompat{\htau}{\htau'}$, and
%     \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
%       $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
%     \end{enumerate}
%   \end{quote}
%   hold, then $\zexp' = \zexp''$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
  \zexp''$.
\end{enumerate}
\end{theorem}

In order to maintain determinism, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can define this notion either direrctly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement. The supplement does the latter. The key rule establishes that arrow types are inconsistent with $\tnum$:
% \begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
% \end{subequations}
% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.

@@

The bidirectional action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
\begin{equation}\label{rule:action-subsume}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}\\\\
    \zexp, \alpha~\mathsf{subsumes}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.  The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e. Rule (\ref{rule:ana-subsume}).

We need to make some exceptions in this rule to maintain determinism, because certain construction actions, described in Sec. \ref{sec:construction}, behave slightly differently depending on whether a type is known. We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
\begin{subequations}
\begin{equation}\label{rule:subsumes}
\inferrule{
  \alpha \neq \aConstruct{\fasc}\\
  (\forall x) \alpha \neq \aConstruct{\flam{x}} 
}{
  \zexp, \alpha~\mathsf{subsumes}
}
\end{equation}
\begin{equation}\label{rule:subsumes}
\inferrule{
  \zexp \neq \zwsel{\hexp}
}{
  \zexp, \alpha~\mathsf{subsumes}
}
\end{equation}
\end{subequations}
Algorithmically, this captures the notion that subsumption should be the case of last resort.

@@

\paragraph{Ascription} Moving on to expression actions, we start to see some more interesting rules. The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type. In the first case, the synthesized type appears in the ascription:
\begin{equation}
  \label{r:constructasc}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}{\htau}
  }
\end{equation}
In the second case, the type provided for analysis appears in the ascription:
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}
  }
\end{equation}
This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).

@@

\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole. If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the focused empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
\begin{equation}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}
Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).

Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
is inserted inside a hole, which defers the type inconsistency as previously discussed:\todo{supp}
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hhole{
        \hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}
      }}
  }
\end{equation}

@@

Witnesses...\todo{talk about this}

>>>>

% The type action judgement appears as a premise in the rules that handle Z-expressions where the cursor is inside a . 
% \[\begin{array}{ll}
% %\textbf{Judgement Form} & \textbf{Description}\\
% & 
% \end{array}\]
% \[
% \begin{array}{l}
% \fbox{$\performTyp{\ztau}{\alpha}{\ztau'}$} ~~ \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
% \fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
% \quad \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
% \fbox{$\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$} \\
% \quad \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
% \quad \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
% \quad \text{against $\htau$}
% \end{array}\]
\subsubsection{Sensibility}

Before giving the rules for these judgements, let us state the key metatheorem: \emph{sensibility}. This metatheorem deeply inform the design of the rules, given starting in Sec. \ref{sec:action-subsumption}. Its proof, which is mechanized in Agda, is by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules. 
\begin{theorem}[Action Sensibility]\label{thrm:actsafe} Both of the following hold:\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ 
   then
  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{enumerate}
\end{theorem}
\noindent In other words, if an edit state (i.e. a Z-expression) is statically meaningful, i.e. its cursor erasure is well-typed, then performing an action on it leaves the resulting edit state statically meaningful. In particular, the first clause of Theorem \ref{thrm:actsafe} establishes that when an action is performed on an edit state whose cursor erasure synthesizes an H-type, the result is an edit state whose cursor erasure also synthesizes some (possibly different) H-type. The second clause establishes that when an action is performed using the analytic action judgement on an edit state whose cursor erasure analyzes against some H-type, the result is a Z-expression whose cursor erasure also analyzes against the same H-type. % Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.

@@

% \subsubsection{Determinism}
% Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
% \begin{theorem}[Action Determinism] All of the following hold:
% \label{thrm:actdet}
% \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
% \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
% \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
%   $\zexp' = \zexp''$ and $\htau' = \htau''$.
% % \item If all of

%   \begin{quote}
%     \begin{enumerate}
%     \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
%     \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
%     \item $\tcompat{\htau}{\htau'}$, and
%     \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
%       $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
%     \end{enumerate}
%   \end{quote}
%   hold, then $\zexp' = \zexp''$.
% % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
% %   \zexp''$.
% % \end{enumerate}
% \end{theorem}

\subsubsection{Type Inconsistency}
In order to define the action semantics, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can define this notion either direrctly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement. The supplement does the latter. The key rule establishes that arrow types are inconsistent with $\tnum$:
% \begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
% \end{subequations}
% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.

@@

The bidirectional action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
\begin{equation}\label{rule:action-subsume}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}%\\\\
    % \zexp, \alpha~\mathsf{subsumes}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.  The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e. Rule (\ref{rule:ana-subsume}).

% We need to make some exceptions in this rule to maintain determinism, because certain construction actions, described in Sec. \ref{sec:construction}, behave slightly differently depending on whether a type is known. We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
% \begin{subequations}
% \begin{equation}\label{rule:subsumes}
% \inferrule{
%   \alpha \neq \aConstruct{\fasc}\\
%   (\forall x) \alpha \neq \aConstruct{\flam{x}} 
% }{
%   \zexp, \alpha~\mathsf{subsumes}
% }
% \end{equation}
% \begin{equation}\label{rule:subsumes}
% \inferrule{
%   \zexp \neq \zwsel{\hexp}
% }{
%   \zexp, \alpha~\mathsf{subsumes}
% }
% \end{equation}
% \end{subequations}
% Algorithmically, this captures the notion that subsumption should be the case of last resort.

@@

\paragraph{Ascription} Moving on to expression actions, we start to see some more interesting rules. The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type. In the first case, the synthesized type appears in the ascription:
\begin{equation}
  \label{r:constructasc}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}{\htau}
  }
\end{equation}
In the second case, the type provided for analysis appears in the ascription:
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}
  }
\end{equation}
% This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).

@@

\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole. If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the focused empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
\begin{equation}\label{rule:performAna-lam-1}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}
% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).

Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
is inserted inside a hole, which defers the type inconsistency as previously discussed:\todo{supp}
\begin{equation}\label{rule:performAna-construct-lam-2}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hhole{
        \hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}
      }}
  }
\end{equation}

@@
